\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis

### 2023-01-24 src/tex/ms.tex
comments throughout

The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
More general examples can lead to a significant increase in dimension.\te{add citations}  \mi{assuming you don't mean TGR, since it's mentioned below, do you mean calibration and tides?: for cal we can cite \cite{Farr:2014,Vitale:2020gvb}, and for tides maybe the BNS properties paper \cite{LIGOScientific:2018hze}, unless you want to cite a bunch of tidal waveform papers.}
% In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
% \kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
% \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
% \AZ{Add PPE and maybe TIGER refs..?}
Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
\kw{Need more reference and some rewording}\te{added refs, happy?}
% All of these tasks require significant computation.
Overall, analyzing GWs therefore requires significant computation.
% \kw{This could be triggering to some people, since population analysis could be quite cheap}
In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.

@@

\AZ{Add refs here:}\te{Added a bunch, let me know if more are needed!}
There are a variety of gravitational waveforms currently used in analysis pipelines.
They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate)~\citep{Varma:2018mmi,Varma:2019csw}.
%Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in the frequency domain using closed-form expressions.
This makes a
% \sout{In particular, their closed form expressions and fixed size frequency grids make a }
\jax implementation that complies with the constraints of JIT compilation simple.
NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity simulations, are in principle also straightfoward to implement in \jax.
EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.][]{Cotesta:2020qhw}.
% to produce a waveform
%require dynamic frequency grids and are therefore tricky to implement in \jax.
%\AZ{What does this mean? I don't understand why EOB needs dynamic frequency grids. We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.} 
%In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
%\AZ{rephrase this in a more positive fashion:}
%NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}

@@

As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to numerical relativity waveforms.
EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
IMRPhenom waveforms are therefore ideally-suited for AD, especially using \jax. 
In this paper we focus on the aligned-spin, circular-orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.

@@

To demonstrate our implementation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
\begin{equation}
    \label{eq:inner_prod}
    \left(h_1|h_2\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h^*_1(f) h_2(f)}{S_n(f)}\, ,
\end{equation}
where $S_n$ is the (one-sided) noise power spectral density (PSD) and \AZ{$h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.}
\sout{ evaluated at the intrinsic parameter point $\theta_1$.}
We can then normalize the inner product through
\begin{equation}
    \left[h_1|h_2\right] = \frac{\left(h_1|h_2\right)}{\sqrt{\left(h_1|h_1\right)\left(h_2|h_2\right)}}\, .
\end{equation}
Now we are ready to define the \textit{match} which is given by 
\begin{equation}
    \mathrm{m}(h_1|h_2) \equiv \max_{\Delta t_c,\, \Delta \phi_c} \left[h_1 \, \middle| \, h_2 \right]\,
\end{equation}
where \AZ{$\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.}

Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
\AZ{Say what PSD was used and reference it.}
This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_1|h_2)$ across the entire parameter space.\footnote{
    Specifically, we use $10^4$ points \AZ{varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$} in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$. In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
}
\AZ{Here, $h_1$ corresponds to the \ripple waveform implementation and $h_2$ is the \lalsuite implementation, evaluated at the same point in parameter space.} 
From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
\AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h2)=0$ causing points to not appear in the log scale colorbar.
% In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
\begin{figure}[t]
    \script{random_matches.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/random_matches.pdf}
        \caption{
            Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
            The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
            It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
            \dfm{I think this figure could use some work. Maybe the points could be smaller to make them easier to resolve. And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch. Right now it looks less impressive!}
        }
        \label{fig:match}
    \end{centering}
\end{figure}

Note that in General Relativity (GR) the total mass $M = m_1 + m_2$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
\sout{the system is simply an overall scaling of the waveforms} 
\sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
\AZ{Revisit this after defining PSD above:} 
This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.

There remains some slight deviation between the two waveform implementations at high total mass. 
This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax. \dfm{What dimension of interpolation do you need? Maybe this is something we could implement...}
Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of $5\times10^5$ values, which we then linearly interpolate during waveform generation.
Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
% The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable for realistic data analysis tasks.

@@

The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases} many more parameters \AZ{can be} added.

@@

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are \sout{commonly
regarded as the most accurate,} 
\AZ{in principle the highest fidelity signal model,}
they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating ``approximants'' of the
full numerical relativity waveforms, which are essentially simpler ansatzes that
\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
covered by numerical simulations, such as extremes of parameter space or earlier 
phases of the binary inspiral}. 
\amc{redundant -- remove or merge with earlier material.}

@@

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients were fitted in segments with amplitude and phase separated.
Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
Each of these segments has their own set of fitting coefficients.
After obtaining the fitting coefficient for individual segments, they are then ``stitched'' together such that they are continuous in the first derivative.
The process of stitching introduces some additional
inaccuracy in the waveform model, as the connections would affect the originally
fitted segments.

@@

\AZ{Maybe comment on optimization method used for EOB waveforms?
I think they do interesting Monte Carlo etc. Check what method is used 
for IMRPhenomXPHM?}

@@

The first step in fitting all coefficients is to define a loss function that
measures the goodness-of-fit of the optimization procedure, which we choose it
to be the mismatch between the NR waveform and the approximant waveform:
\begin{equation}
    \label{eq:mismatch}
	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda)),
\end{equation}
where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
generated by the numerical relativity simulation. 
Given the loss function, we use gradient descent to find the optimal fitting
coefficients.
\begin{equation}
	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}, 
\end{equation}
where $\alpha$ is the learning rate. We set $\alpha$ to be a very small number, i.e. $10^{-6}$, since the initial guess already lies close to the desired optimum. We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.  \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
\amc{say something to explain why this is working. I.e.: does the loss initially drop and then plateau?}
% Add description of the optimization procedure.

As a demonstration, Fig. \ref{fig:loss_compare} shows the relative error of
original and optimized waveform at different frequencies. In this optimization
procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
as training waveforms to fit waveform coefficients. \amc{how many are there?}
We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
In particular, the error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy. \amc{so this is a waveform in the training set? And I don't follow exactly what you did. Is this just fitting a single waveform?}

We can generalize the fitting procedure described above to a collection of waveforms.
In that case, the loss function is defined as the average of the mismatch of
individual waveforms, given by
\begin{equation}
	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i,
    \label{eq:multi_loss}
\end{equation}
where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$
is the total number of training waveforms used in optimization. This
optimization is more difficult since we are applying the same set of
coefficients to waveforms with different intrinsic parameters, such as mass
ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
waveforms with different intrinsic parameters implies there are trade-offs in
performance in different regions of the intrinsic parameter space.
\amc{this means your fit generally will depend on how the training waveforms are distributed over parameter space.}

@@

In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches.
One can see the distribution of mismatch after optimization has smaller mismatch compared to before optimization,
showing the optimization improves the accuracy of the waveform model.
% However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
While many waveforms improved their accuracy, some retain similar mismatches. 
\AZ{I don't understand this claim. The spin degeneracy is present in the physics of the waveform; $\chi_{\rm eff}$ is just one way of parametrizing the spins that tries to reduce the degeneracy.}
This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. 
In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
Nevertheless, the peak of the original waveform distribution has moved to the left end of the distribution, indicating that \sout{the} \AZ{our AD-assisted} optimization procedure \AZ{provides an improved implementation of the model}. 
\sout{Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses. }

@@

Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.}
Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
Although in practice this assumption is often violated for realistic detector noise, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.

>>>>

The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
More general examples can lead to a significant increase in dimension.\te{add citations}  \mi{assuming you don't mean TGR, since it's mentioned below, do you mean calibration and tides?: for cal we can cite \cite{Farr:2014,Vitale:2020gvb}, and for tides maybe the BNS properties paper \cite{LIGOScientific:2018hze}, unless you want to cite a bunch of tidal waveform papers.}
% In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
% \kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
% \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}.
% \AZ{Add PPE and maybe TIGER refs..?}
Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
\kw{Need more reference and some rewording}\te{added refs, happy?}
% All of these tasks require significant computation.
Overall, analyzing GWs therefore requires significant computation.
% \kw{This could be triggering to some people, since population analysis could be quite cheap}
In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.

@@

\AZ{Add refs here:}\te{Added a bunch, let me know if more are needed!}
There are a variety of gravitational waveforms currently used in analysis pipelines.
They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom) \citep{Husa:2015iqa,Khan:2015jqa,Hannam:2013oca,Pratten:2020ceb}, and numerical relativity surrogate (NRsurrogate)~\citep{Blackman:2017pcm,Varma:2018mmi,Varma:2019csw}.
%Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in the frequency domain using closed-form expressions.
This makes a
% \sout{In particular, their closed form expressions and fixed size frequency grids make a }
\jax implementation that complies with the constraints of JIT compilation simple.
NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity simulations, are in principle also straightfoward to implement in \jax.
EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.][]{Cotesta:2020qhw}.
% to produce a waveform
%require dynamic frequency grids and are therefore tricky to implement in \jax.
%\AZ{What does this mean? I don't understand why EOB needs dynamic frequency grids. We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.} 
%In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
%\AZ{rephrase this in a more positive fashion:}
%NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}

@@

As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to numerical relativity waveforms.
\mi{I think this is weird to say: the surrogate is fully trained on (``calibrated to'') NR}
EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
IMRPhenom waveforms are therefore ideally-suited for AD, especially using \jax. 
In this paper we focus on the aligned-spin, circular-orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.

@@

To demonstrate our implementation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
\begin{equation}
    \label{eq:inner_prod}
    \left(h_1|h_2\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h^*_1(f) h_2(f)}{S_n(f)}\, ,
\end{equation}
where $S_n$ is the (one-sided) noise power spectral density (PSD) and \AZ{$h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.}
\sout{ evaluated at the intrinsic parameter point $\theta_1$.}
We can then normalize the inner product through
\begin{equation}
    \left[h_1|h_2\right] = \frac{\left(h_1|h_2\right)}{\sqrt{\left(h_1|h_1\right)\left(h_2|h_2\right)}}\, .
\end{equation}
Now we are ready to define the \textit{match} which is given by 
\begin{equation}
    \mathrm{m}(h_1|h_2) \equiv \max_{\Delta t_c,\, \Delta \phi_c} \left[h_1 \, \middle| \, h_2 \right]\,
\end{equation}
where \AZ{$\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.}
\mi{I think it'd be good to define the mismatch $\mathcal{M} \equiv 1- \mathrm{m}$, since it's the standard quantity to discuss, and it's what we plot}

Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
\AZ{Say what PSD was used and reference it.}
This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_1|h_2)$ across the entire parameter space.\footnote{
    Specifically, we use $10^4$ points \AZ{varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$} in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$. In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
}
\AZ{Here, $h_1$ corresponds to the \ripple waveform implementation and $h_2$ is the \lalsuite implementation, evaluated at the same point in parameter space.} 
From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
\AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h2)=0$ causing points to not appear in the log scale colorbar.
% In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
\mi{to Aaron's point, maybe say ``0 up to numerical precision''?}

\begin{figure}[t]
    \script{random_matches.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/random_matches.pdf}
        \caption{
            Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
            The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
            It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
            \dfm{I think this figure could use some work. Maybe the points could be smaller to make them easier to resolve. And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch. Right now it looks less impressive!}
            \mi{I agree. Furthermore, it looks like there's a subpopulation of darker points hiding under the brighter ones up to and around $M \sim 100$? is there a dependence on the mass ratio? it'd be good to comment on this, and maybe show a corner plot adding $q$ to the mix. Also, $M$ is lacking units (I assume $M_\odot$, but it should be shown in the label)}
        }
        \label{fig:match}
    \end{centering}
\end{figure}

Note that in General Relativity (GR) the total mass $M = m_1 + m_2$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
\sout{the system is simply an overall scaling of the waveforms} 
\sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
\AZ{Revisit this after defining PSD above:} 
This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
\mi{it's important to say what PSD was used}

There remains some slight deviation between the two waveform implementations at high total mass. 
This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax. \dfm{What dimension of interpolation do you need? Maybe this is something we could implement...}
Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of $5\times10^5$ values, which we then linearly interpolate during waveform generation.
Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
% The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable effect for realistic data analysis tasks.

@@

The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension added, $D$, one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases} many more parameters \AZ{can be} added.

@@

\mi{this paragraph is redundant, I'd remove it}
Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are \sout{commonly
regarded as the most accurate,} 
\AZ{in principle the highest fidelity signal model,}
they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating ``approximants'' of the
full numerical relativity waveforms, which are essentially simpler ansatzes that
\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
covered by numerical simulations, such as extremes of parameter space or earlier 
phases of the binary inspiral}. 
\amc{redundant -- remove or merge with earlier material.}

@@

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients were fitted in segments with amplitude and phase separated.
\mi{this previous sentence sounds odd: do we mean that amplitude and phase were fitted independently?}
Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
Each of these segments has their own set of fitting coefficients.
After obtaining the fitting coefficient for individual segments, they are then ``stitched'' together such that they are continuous in the first derivative.
The process of stitching introduces some additional
inaccuracy in the waveform model, as the connections would affect the originally
fitted segments.

@@

\AZ{Maybe comment on optimization method used for EOB waveforms?
I think they do interesting Monte Carlo etc. Check what method is used 
for IMRPhenomXPHM?}
\mi{I agree. it'd be good to comment on whether this is still done like this in
modern waveforms.}

@@

The first step in fitting all coefficients is to define a loss function that
measures the goodness-of-fit of the optimization procedure, which we choose it
to be the mismatch between the NR waveform and the approximant waveform:
\begin{equation}
    \label{eq:mismatch}
	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda)),
\end{equation}
where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
generated by the numerical relativity simulation. 
\mi{I think the mismatch should be defined above, right after the match is defined. We should also state what PSD and sampling parameters are used}
Given the loss function, we use gradient descent to find the optimal fitting
coefficients.
\begin{equation}
	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}, 
\end{equation}
where $\alpha$ is the learning rate. We set $\alpha$ to be a very small number, i.e. $10^{-6}$, since the initial guess already lies close to the desired optimum. We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.  \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
\amc{say something to explain why this is working. I.e.: does the loss initially drop and then plateau?}
% Add description of the optimization procedure.

As a demonstration, Fig. \ref{fig:loss_compare} shows the relative error of
the original and optimized waveforms at different frequencies \mi{for a particular set of parameters?}. In this optimization
procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
as training data to fit waveform coefficients. \amc{how many are there?}
We can see the error of the optimized waveform is lower than that of the original waveform for most of the domain.
In particular, the error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy. \amc{so this is a waveform in the training set? And I don't follow exactly what you did. Is this just fitting a single waveform?}
\mi{I was also confused by this}

We can generalize the fitting procedure described above to a collection of waveforms.
In that case, the loss function is defined as the average of the mismatch of
individual waveforms, given by
\begin{equation}
	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i,
    \label{eq:multi_loss}
\end{equation}
where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$
is the total number of training waveforms used in optimization. This
optimization is more difficult since we are applying the same set of
coefficients to waveforms with different intrinsic parameters, such as mass
ratio and spins. From Eq.~\eqref{eq:multi_loss}, we can see the averaging between
waveforms with different intrinsic parameters implies there are trade-offs in
performance in different regions of the intrinsic parameter space.
\amc{this means your fit generally will depend on how the training waveforms are distributed over parameter space.}

@@

In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches.
\mi{for all parameters in the training set?}
One can see the distribution of mismatch \mi{mean?} after optimization has smaller mismatch compared to before optimization,
showing the optimization improves the accuracy of the waveform model \mi{on average?}.
% However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
While many waveforms improved their accuracy, some retain similar mismatches \mi{``waveforms'' is overused and obfuscates this discussion: it's indistinctly used to refer to the NR data, the approximants in general, and specific sets of parameters---I think some rewriting is needed, e.g., here we could say ``While the mismatch generally improves for several parameter choices after training, it remains the same for others.''}. 
\AZ{I don't understand this claim. The spin degeneracy is present in the physics of the waveform; $\chi_{\rm eff}$ is just one way of parametrizing the spins that tries to reduce the degeneracy.}
This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. 
In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
Nevertheless, the peak of the original waveform distribution has moved to the left end of the distribution, indicating that \sout{the} \AZ{our AD-assisted} optimization procedure \AZ{provides an improved implementation of the model}. 
\sout{Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses. }

@@

Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.} \mi{routine?}
Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters could be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
Although in practice this assumption is often violated for realistic detector noise, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.
